Lessons learned from bugs in models of human history

### 2020-08-03 paper.tex
some initial changes to text to start addressing reviewer comments

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{fullpage}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
\AfterEndEnvironment{minted}{\end{tcolorbox}}%
\usepackage{url}

@@

\title{Lessons learned from bugs in models of human history}
\author[1]{Aaron P. Ragsdale}
\author[1]{Dominic Nelson}
\author[1,$\star$]{Simon Gravel}
\author[2,$\star$]{Jerome Kelleher}
\affil[1]{McGill Genome Centre and Department of Human Genetics,
McGill University, Montr\'{e}al, Qu\'{e}bec, Canada}
\affil[2]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
University of Oxford, Oxford, United Kingdom}
\affil[$\star$]{Joint senior authors, listed alphabetically}
\maketitle

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
Despite these initial checks being made, the simulation
was performed without passing the parameter defining demographic events,
so that the three populations
never merged and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: heterozygosity was more than
doubled in African populations and reduced by more than half in Eurasian populations
relative to the correct model.

@@

Firstly, if possible, we recommend avoiding the implementation of
demographic models entirely by using \stdpopsim~\citep{adrion2019community}.
This is a ``standard library" of quality-controlled
models and simulation resources for a growing number of commonly studied species.
The resource is built around an open-source community development model
with rigorous code review and quality control procedures.
For example, when a contributor adds a new model it is not fully integrated
into the catalog until a second, entirely independent, implementation of the
model is provided by another developer. If the original and ``QC" versions
of the model are precisely equal, then we can have a reasonable degree
of confidence in correctness. It is through this QC process that we discovered
the misspecified model in \msprime's documentation.
If the required model is not present in \stdpopsim, then particular care should
be taken to validate the implementation.
This could include verification through code review or comparison with an
independent implementation of the model. If the model is of wider interest to
the community, we would encourage developers to contribute it to \stdpopsim:
at the very least, the quality-control procedures in place provide a strong
reassurance of correctness.

Secondly, regardless of whether the model has been implemented
locally or obtained from a resource like \stdpopsim,
basic statistical validation should \emph{always} be performed on the
results. Errors are all too easy to make, and an analysis of the
basic statistical properties of the simulations is essential
due diligence. For example, a brief analysis of $F_{ST}$ values
would likely would have prevented the error in~\citet{martin2017human}.
It is understandable that such analyses were not undertaken in this
case, as processing such a huge dataset (600,000 samples) was
highly challenging. Recent progress has made statistical analysis this scale much
easier~\citep{ralph2020efficiently}, but the problems would have been
apparent if smaller pilot simulations were undertaken and analysed.
Graphical inspection of demographic models can also help identify issues, especially if
visualization can be automated. The demography debugging tool in \msprime\
summarizes demographic events occurring over time, and various efforts are underway to facilitate
the specification and visualization of population genetics models, such as~\citet{zhou2018popdemog}
and the \texttt{demography} package used in this paper.

@@

\bibliographystyle{plainnat}
\bibliography{paper}

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
\caption{\textbf{The transferability of PRS under neutrality}.
    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
    events in the past, so continental populations were simulated as isolated with low levels of migration
    for all time. The correct model has historical events as shown in Fig.~\ref{fig:ooa_stats}A.
    \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct
    demographic model.
    GWAS summary statistics were computed from 10,000 case and control subjects in the European
    population, and the distribution of inferred PRS were compared across the three simulated populations.
    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
    Unlike the original study, we do not observe large differences in mean inferred PRS across
    the three populations (C, D), and while risk prediction in the African and East Asian populations is
    still reduced compared to the European population, the reduction in prediction accuracy is not as
    large as reported in the original study (E-G).
    Panel D shows a comparison of PRS distributions for a simulation replicate with 1,000 causal variants.
    For each population in the violin plots, the original correlations from~\citet{martin2017human}
    are shown on the left, and correlations using the correct model are shown on the right.
    For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.
}
\label{fig:prs}
\end{center}
\end{figure}

>>>>

\documentclass{article}
\usepackage[super,comma]{natbib}
\usepackage{fullpage}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
\AfterEndEnvironment{minted}{\end{tcolorbox}}%
\usepackage{url}

@@

\title{Lessons learned from bugs in models of human history}
\author[1]{Aaron P. Ragsdale}
\author[1]{Dominic Nelson}
\author[1,$\star$]{Simon Gravel}
\author[2,$\star$]{Jerome Kelleher}
\affil[1]{McGill Genome Centre and Department of Human Genetics,
McGill University, Montr\'{e}al, Qu\'{e}bec, H2X 3C9, Canada}
\affil[2]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
University of Oxford, Oxford, \aprcomment{postal code} United Kingdom}
\affil[$\star$]{Joint senior authors, listed alphabetically}
\maketitle

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
Despite these initial checks being made, the simulation
was performed without passing the list of demographic events such as historical
changes in size or population splits, so that the three populations
never merged and remained separated with low levels of migration (Fig~\ref{fig:prs}A).
This leads to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: heterozygosity was more than
doubled in African populations and reduced by more than half in Eurasian populations
relative to the correct model.

@@

Firstly, if possible, we recommend using the \stdpopsim~\citep{adrion2019community}
library, which removes the need to re-implement demographic models from scratch.
This is a ``standard library" of quality-controlled
models and simulation resources for a growing number of commonly studied species.
The resource is built around an open-source community development model
with rigorous code review and quality control procedures.
For example, when a contributor adds a new model it is not fully integrated
into the catalog until a second, entirely independent, implementation of the
model is provided by another developer. If the original and ``QC" versions
of the model are precisely equal, then we can have a reasonable degree
of confidence in correctness. It is through this QC process that we discovered
the misspecified model in \msprime's documentation.
If the required model is not present in \stdpopsim, then particular care should
be taken to validate the implementation.
This could include verification through code review or comparison with an
independent implementation of the model. If the model is of wider interest to
the community, we would encourage developers to contribute it to \stdpopsim:
at the very least, the quality-control procedures in place provide a strong
reassurance of correctness.

Secondly, regardless of whether the model has been implemented
locally or obtained from a resource like \stdpopsim,
basic statistical validation should \emph{always} be performed on the
results. Errors are all too easy to make, and an analysis of the
basic statistical properties of the simulations is essential
due diligence. For example, a brief analysis of $F_{ST}$ values
would likely would have prevented the error in~\citet{martin2017human}.
It is understandable that such analyses were not undertaken in this
case, as processing such a huge dataset (600,000 samples) was
highly challenging. Recent progress has made statistical analysis scale much
easier for large sample sizes~\citep{ralph2020efficiently}, but the problems would have been
apparent if smaller pilot simulations were undertaken and analysed.
Graphical inspection of demographic models can also help identify issues, especially if
visualization can be automated. The demography debugging tool in \msprime\
summarizes demographic events occurring over time, and various efforts are underway to facilitate
the specification and visualization of population genetics models, such as~\citet{zhou2018popdemog}
and the \texttt{demography} package used in this paper.

@@

\bibliographystyle{unsrtnat}
\bibliography{paper}

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
\caption{\textbf{The transferability of PRS under neutrality}.
    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
    events in the past, so continental populations were simulated as isolated with low levels of migration
    for all time. The correct model has historical events as shown in Fig.~\ref{fig:ooa_stats}A.
    \textbf{(B-G)} We repeated the simulation experiment in~\citet{martin2017human} using the correct
    demographic model.
    GWAS summary statistics were computed from 10,000 case and control subjects in the European
    population, and the distribution of inferred PRS were compared across the three simulated populations.
    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
    Unlike the original study, we do not observe large differences in mean inferred PRS across
    the three populations (C, D), and while risk prediction in the African and East Asian populations is
    still reduced compared to the European population, the reduction in prediction accuracy is not as
    large as reported in the original study (E-G).
    Panel D shows a comparison of PRS distributions for a simulation replicate with 1,000 causal variants.
    For each population in the violin plots, the original correlations from~\citet{martin2017human}
    are shown on the left, and correlations using the correct model are shown on the right.
    For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.
}
\label{fig:prs}
\end{center}
\end{figure}
